Location and Scale-Invariant Power Transformations for Transforming Data to

### 2025-04-09 manuscript_rmarkdown/manuscript.Rmd
Changed text to definitions.

---
title: "Location and Scale-Invariant Power Transformations for Transforming Data to
  Normality"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{amsthm}
- \usepackage{booktabs}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

Many statistical and some machine learning methods assume normality of the underlying data, e.g. analysis of variance and linear discriminant analysis.
However, numerical features in datasets may strongly deviate from normal distributions, e.g. by being skewed.
Power transformations aim to stabilise variance and improve normality of such features [@Bartlett1947-rx; @Tukey1957-rt].
The two most commonly used transformations are that of @Box1964-mz and @Yeo2000-vw.
The Box-Cox transformation of a feature value $x_i$ of feature $\mathbf{X}=\left\{x_1, x_2, \ldots, x_n \right\}$ under the transformation parameter $\lambda$ is defined as:

```{=latex}
\begin{equation}
\label{eqn:box-cox-original}
\phi_{\text{BC}}^\lambda (x_i) = 
\begin{cases}
\left(x_i^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i) & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
One limitation of the Box-Cox transformation is that it is only defined for $x_i > 0$. In contrast, the Yeo-Johnson transformation under the transformation parameter $\lambda$ is defined for any $x_i \in \mathbb{R}$:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}
```
The $\lambda$-parameter is typically optimised using maximum likelihood estimation under the assumption that the transformed feature is normally distributed.
As noted by Raymaekers and Rousseeuw, this approach is sensitive to outliers, and robust versions of Box-Cox and Yeo-Johnson transformations were devised [@Raymaekers2024-zf].

@@

Box-Cox and Yeo-Johnson transformations are modified by introducing shift parameter $x_0$ and scale parameter $s$ into equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}.
The location- and scale-invariant Box-Cox transformation of a feature value $x_i$ of feature $\mathbf{X}$ under transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$ is then:

```{=latex}
\begin{equation}
\label{eqn:box-cox-invariant}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
where $x_i - x_0 > 0$. Likewise, the location- and scale-invariant Yeo-Johnson transformation of a feature value $x_i$ under transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$ is:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-invariant}
\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left( 1 + \frac{x_i - x_0}{s}\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log\left[1 + \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - \frac{x_i - x_0}{s}\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0 < 0\\
-\log\left[1 - \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}
```

@@

```{=latex}
\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{BC}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
subject to $x_i - x_0 > 0$.
$\mu$ and $\sigma^2$ are the mean and variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.
Similarly, the location- and scale-invariant Yeo-Johnson log-likelihood function is:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{YJ}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\mu$ and $\sigma^2$ are the mean and variance of the Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.

@@

\begin{definion}[Asymmetric generalised normal distribution probability density function]
Let $\alpha \in (0,1)$ be a skewness parameter.
Let $\mu \in \mathcal{R}$ be a location parameter.
Let $\sigma > 0$ be a shape parameter.

@@

\begin{definition}[central portion of a sequence]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathbb{R}\right\}$ be a finite sequence of length $n > 0$,
ordered so that $x_1 \leq x_2 \leq \ldots \leq x_n$.
Let $p_i = \frac{i - 1/3}{n + 1/3}$, with $i = 1, 2, \ldots n$ be the percentile value
corresponding to each element in $\mathbf{X}$. For elements with tied values in $\mathbf{X}$,
percentile values are replaced by the average in their group, i.e.
if $x_j = x_{j+1} = \ldots = x_{j+m}$, then 
$p_j' = p_{j+1}' = \ldots = p_{j+m}' = 1/(m+1)\sum_j^{j+m}p_j$.
Furthermore, let central portion $\kappa \in (0,1)$.

@@

To assess whether the proposed power transformations lead to values of $\lambda$ 
that are invariant to location and scale of the distribution, we simulated three different sequences
We first randomly drew $10000$ values from a normal distribution: 
$\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, 
or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution was a right-skewed generalised normal distribution 
$\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution was a left-skewed generalised normal distribution 
$\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then computed transformation parameter $\lambda$ using the original definitions (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the location- and scale-invariant definitions (Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) for each distribution.
To assess location invariance, a positive value $d_{\text{shift}}$ was added to each distribution with $d_{\text{shift}} \in [1, 10^6]$.
Similarly, to assess scale invariance, each distribution was multiplied by a positive value $d_{\text{scale}}$, where $d_{\text{scale}} \in [1, 10^6]$.

>>>>

---
title: "Location and Scale-Invariant Power Transformations for Transforming Data to
  Normality"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{amsthm}
- \usepackage{booktabs}
- \newtheorem*{definition}{Definition}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

Many statistical and some machine learning methods assume normality of the underlying data,
e.g. analysis of variance and linear discriminant analysis.
However, numerical features in datasets may strongly deviate from normal distributions, e.g. by being skewed.
Power transformations aim to stabilise variance and improve normality of such features [@Bartlett1947-rx; @Tukey1957-rt].
The two most commonly used transformations are that of @Box1964-mz and @Yeo2000-vw.

The Box-Cox transformation is defined as:

\begin{definition}[Box-Cox power transformation]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i > 0 \right\}$ be a finite sequence of length $n > 0$.
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
Then the Box-Cox power transformation of element $x_i$ is defined as:
\begin{equation}
\label{eqn:box-cox-original}
\phi_{\text{BC}}^\lambda (x_i) = 
\begin{cases}
\left(x_i^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i) & \text{if } \lambda = 0
\end{cases}
\end{equation}
\end{definition}

One limitation of the Box-Cox transformation is that it is only defined for $x_i > 0$. 
In contrast, the Yeo-Johnson transformation under the transformation parameter $\lambda$ is defined for any $x_i \in \mathbb{R}$, as follows:

\begin{definition}[Yeo-Johnson power transformation]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathbb{R} \right\}$ be a finite sequence of length $n > 0$.
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}
\end{definition}

The $\lambda$-parameter is typically optimised using maximum likelihood estimation under the assumption that the transformed feature is normally distributed.
As noted by Raymaekers and Rousseeuw, this approach is sensitive to outliers, 
and robust versions of Box-Cox and Yeo-Johnson transformations were devised [@Raymaekers2024-zf].

@@

Box-Cox and Yeo-Johnson transformations are modified by introducing
shift parameter $x_0$ and scale parameter $s$ into equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}.
The location- and scale-invariant Box-Cox transformation of a feature value 
$x_i$ of feature $\mathbf{X}$ under transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$ 
is then defined as:

\begin{definition}[Location- and scale- invariant Box-Cox power transformation]
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
Let $x_0 \in \mathcal{R}$ and $s > 0$ be location and scale parameters.
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i - x_0 > 0 \right\}$ be a finite sequence of length $n > 0$.

Then the location- and scale-invariant Box-Cox power transformation of element $x_i$ is defined as:

\begin{equation}
\label{eqn:box-cox-invariant}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}
\end{definition}

Likewise, the location- and scale-invariant Yeo-Johnson transformation of a feature value $x_i$ 
under transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$ is defined as:

\begin{definition}[Location- and scale- invariant Yeo-Johnson power transformation]
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
Let $x_0 \in \mathcal{R}$ and $s > 0$ be location and scale parameters.
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathcal{R} \right\}$ be a finite sequence of length $n > 0$.

Then the location- and scale-invariant Yeo-Johnson power transformation of element $x_i$ is defined as:

\begin{equation}
\label{eqn:yeo-johnson-invariant}
\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left( 1 + \frac{x_i - x_0}{s}\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log\left[1 + \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - \frac{x_i - x_0}{s}\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0 < 0\\
-\log\left[1 - \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}
\end{definition}

@@

\begin{definition} [Log-likelihood function of the location- and scale- invariant Box-Cox power transformation]
Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:box-cox-invariant}).
Let $\mu$ and $\sigma^2$ be the mean and variance of the transformed sequence $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.

Then the log-likelihood function of the location- and scale- invariant Box-Cox power transformation is defined as:
\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{BC}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
\end{definition}

Similarly, the location- and scale-invariant Yeo-Johnson log-likelihood function is:

\begin{definition}[Log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation]
Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:yeo-johnson-invariant}).
Let $\mu$ and $\sigma^2$ be the mean and variance of the transformed sequence $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.

Then the log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation is defined as:
\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{YJ}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
\end{definition}

@@

\begin{definition}[Asymmetric generalised normal distribution probability density function]
Let $\alpha \in (0,1)$ be a skewness parameter.
Let $\mu \in \mathcal{R}$ be a location parameter.
Let $\sigma > 0$ be a shape parameter.

@@

\begin{definition}[Central portion of a sequence]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathbb{R}\right\}$ be a finite sequence of length $n > 0$,
ordered so that $x_1 \leq x_2 \leq \ldots \leq x_n$.
Let $p_i = \frac{i - 1/3}{n + 1/3}$, with $i = 1, 2, \ldots n$ be the percentile value
corresponding to each element in $\mathbf{X}$. For elements with tied values in $\mathbf{X}$,
percentile values are replaced by the average in their group, i.e.
if $x_j = x_{j+1} = \ldots = x_{j+m}$, then 
$p_j' = p_{j+1}' = \ldots = p_{j+m}' = 1/(m+1)\sum_j^{j+m}p_j$.
Furthermore, let central portion $\kappa \in (0,1)$.

@@

To assess whether the proposed power transformations lead to values of $\lambda$ 
that are invariant to location and scale of the distribution, we simulated three different sequences.
We first randomly drew $10000$ values from a normal distribution: 
$\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, 
or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution was a right-skewed generalised normal distribution 
$\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution was a left-skewed generalised normal distribution 
$\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then computed transformation parameter $\lambda$ using the original definitions (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the location- and scale-invariant definitions (Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) for each distribution.
To assess location invariance, a positive value $d_{\text{shift}}$ was added to each distribution with $d_{\text{shift}} \in [1, 10^6]$.
Similarly, to assess scale invariance, each distribution was multiplied by a positive value $d_{\text{scale}}$, where $d_{\text{scale}} \in [1, 10^6]$.
